\texttt{aeons}: approximating the end of nested sampling

### 2023-11-27 paper/paper.tex
Further updates

\begin{abstract}
    This paper presents analytic results on the anatomy of nested sampling, from which a technique is developed to estimate the run-time of the algorithm that works for any nested sampling implementation. At each iteration, the dimensionality of the current set of samples is inferred, then used to extrapolate the known likelihood profile using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be estimated. We test these methods on both toy models and true cosmological nested sampling runs. The method finds the right order of magnitude at all times, obtaining the true endpoint within standard error generally around the halfway point, and converges to the true value by the end of the run.  
\end{abstract}

@@

\subsubsection*{Comparisons}
\cref{fig:beta_comparison} shows the three temperatures as a function of compression for two cases, one containing a phase transition and one without. They are consistent in both cases when there is a single dominant phase, but differ during a phase transition. The canonical temperature is the only one that rises monotonically with compression.
\begin{figure}
\begin{center}
    \includegraphics{figures/beta_comparison.pdf}
\end{center}
\caption{Inferred temperatures using the microcanonical, canonical and Bayesian definitions. The shaded regions show the $1-2\sigma$ uncertainties. All are consistent for a single phase, but differ during a phase transition.}
\label{fig:beta_comparison}
\end{figure}
\par
One should keep in mind that despite the above theoretical reasoning, our introduction of the likelihood transformation was ultimately motivated by our wish to utilise the extra degree of freedom it provides. As we will see below, we recommend choosing the exact definition depending on what is useful for the transformation to do.

@@

\section{Endpoint prediction}\label{sec:endpoint}
As described in \citet{supernest} and further explored in the talk and upcoming paper \citet{kcl_talk, scaling_frontier}, the time complexity of nested sampling is
\begin{equation}
    T \propto \langle \nlive \rangle \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
\end{equation}
The second term is the relatively constant time per likelihood evaluation. The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude. The primary unknown during a run, and therefore our primary interest, is the final term: the compression required to get from prior to posterior.

\subsection{The termination prior volume}
To be exact, one wishes to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}). The problem is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is quite far from the posterior bulk. 
\par
In order to get an idea of where the true posterior bulk sits, we need to predict what the posterior looks like past the highest live point. We do this by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points. 
\par
One would never use this predicted posterior to do inference, since more accuracy can always be achieved by simply finishing the run. However, it is more than sufficient for making a prediction for $\log X_\mathrm{f}$. Quantitatively, this proceeds as follows: fit a function $f(X, \phi)$  with some parameters $\phi$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
\begin{equation}
	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
\end{equation}
\begin{equation}\label{endpoint}
	\int_0^{X_\mathrm{f}} f(X, \phi)\ \mathrm{d}X = \epsilon \left( \int_0^{X_i} f(X, \phi)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_\mathrm{f}$ can then be identified by solving the above equation either analytically or numerically. 
\par
Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule. The conversion is easiest in the constant $\nlive$ case; at each iteration $\log X$ decreases by $1/n$, so the total number of iterations $N_\mathrm{f}$ will be
\begin{equation}
	N_\mathrm{f} = - n\log X_\mathrm{f} .
\end{equation}

\subsection{How to extrapolate?}
A key observation is that the Bayesian model dimensionality is the equivalent dimension of the posterior if it were actually Gaussian. Fitting a Gaussian of this dimension to the likelihood profile therefore makes a reasonable approximation to the true distribution, without explicitly assuming the form of the likelihood function. The parameterisation of the Gaussian that we fit is the same as that given in \cref{sec:logL}, which we shall repeat here for clarity;
\begin{equation}\label{gaussian}
    f(X; \phi) = \logLmax - X^{2/d}/2\sigma^2
\end{equation}
The extrapolation then proceeds thus:
\begin{enumerate}[leftmargin=*]
    \item Find the current dimensionality ${d}^{(i)}_G$ of the posterior at the Bayesian temperature
    \item Take the live point profile and do a least squares fit to \eqref{gaussian}, stipulating that $d = {d}^{(i)}_G$ to infer $\logLmax$ and $\sigma$ 
    \item Use the likelihood predicted by these parameters to solve \eqref{endpoint} for $X_\mathrm{f}$
\end{enumerate}
The advantage of fitting a Gaussian is that the procedure can be sped up analytically. Firstly, the least squares regression is trivial because analytic estimators exist; the cost function 
\begin{equation}\label{chi squared}
	C^2(\logLmax, \sigma) = \sum_i \left| \log \Like_i - f(X_i; \logLmax, \sigma) \right| ^2
\end{equation}
is minimised with respect to $(\logLmax, \sigma)$ when
\begin{equation}\label{eq:sigma}
    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i X_i^{2/d}\log \Like_i },
\end{equation}
and
\begin{equation}\label{eq:logLmax}
    \logLmax = \frac{1}{N} \sum_i \log \mathcal{L}_i + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d}.
\end{equation}
Secondly, the termination prior volume can also be obtained analytically. Rewriting \cref{endpoint} in terms of the Gaussian parameters gives
\begin{equation}
	\epsilon = \frac{\int_0^{X_\mathrm{f}} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead}}.
\end{equation}
The integrals have the analytic solution
\begin{equation}
	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X = \frac{d}{2} \cdot \left(\sqrt{2}\sigma\right)^d \cdot \gamma_k
\end{equation}
where $\gamma_k = \Gamma_{d/2}\left(X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
\begin{equation}
    \log X_\mathrm{f} = \frac{d}{2}\log 2\sigma^2	+ \log \Gamma^{-1}_{d/2} \left(\epsilon \gamma_i+ \frac{\epsilon\mathcal{Z}_\mathrm{dead}}{ \left( 2\sigma^2 \right)^{d/2}\Like_\mathrm{max}}\right),\label{eq:xf}
\end{equation}
and $N_\mathrm{f}$ is of course just $-n$ multiplied by this. Intuitively, the above procedure can be thought of as inferring the number of constrained parameters, then extrapolating them up to find the point at which they will be fully constrained. 
\par
One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown it to be far less stable.

\subsection{Alternative approaches}
More comprehensive Bayesian approaches, perhaps including \textit{a priori} information about the likelihood or greater flexibility in the fitting function, could certainly do better than what we have just presented. However, such methods would not befit \textit{run-time} prediction which has a much more limited computational budget, hence the more practical approach we have adopted. Here, we discuss as a benchmark alternative approaches to endpoint estimation that have a comparable computational complexity. 
\subsubsection{Integral progress}
One alternative approach, used in \textsc{Ultranest} \citep{ultranest}, derives a progress bar based on the fraction of the accumulated integral compared to the remaining integral, approximated as 
\begin{equation}\label{eq:integral_progress}
    \mathcal{Z}_\mathrm{rem} \approx \Like_\mathrm{max}^{\mathrm{live}} X_i.
\end{equation}
There are a few problems here. Firstly, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width. Secondly, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is simply be zero for most of the run. Furthermore, approximation \eqref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum.
\par
This approach can be useful in low dimensions when the live points are always near the maximum, but in general is less reliable.

\subsubsection{Extrapolating evidence increments}
Seasoned users of nested sampling might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges. We do this for a spherical Gaussian and compare it to our method. At an intermediate stage of the run, the most recent outputs might look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate}. Extrapolating those data to a linear and exponential profile yields endpoint estimates plotted in the graph to the right. 
\par
The linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile. The increments are also not quite exponential, since the exponential fit leads to a large over-prediction. The predicted endpoint over the course of a run for $d = 16$, $\sigma = 0.01$, as shown in \cref{fig:inc_predictions}, shows the same result. One might expect an average to be more accurate, but this tends to be biased towards the exponential prediction, and there is no obvious choice of weighting that would fix this.
\par
More importantly, we find that for real likelihoods which have an element of noise the extrapolation often diverges, for instance when the increments do not monotonically decrease. Directly extrapolating the evidence increments is therefore far less stable than the previous method, and generally not a reliable method for prediction.
\begin{figure}
\begin{adjustbox}{valign=t, scale=0.72}
\subfloat{\begin{tabular}{|c|c|c|}
\hline
iteration & log Z  & $\Delta\log Z$   \\
\hline
5000 & -1435.8 & 190.8 \\
5500 & -1264.6 & 171.2 \\
6000 & -1123.7 & 140.9 \\
6500 & -991.5 & 132.2 \\
7000 & -885.0 & 106.6 \\
7500 & -790.3 & 94.7 \\
8000 & -702.6 & 87.7 \\
8500 & -619.7 & 82.9 \\
9000 & -551.8 & 67.9 \\
9500 & -492.7 & 59.1 \\
\hline
\end{tabular}}
\end{adjustbox}
\quad
\begin{adjustbox}{valign=t}
\subfloat{\includegraphics{figures/inc_extrapolate.pdf}}
\end{adjustbox}
\caption{Extrapolating the increments of evidence. The left column shows the output of a nested sampling run, and the right column shows the extrapolation.}
\label{fig:inc_extrapolate}
\end{figure}
\begin{figure}
\includegraphics{figures/inc_predictions.pdf}
\caption{Endpoint predictions for a spherical Gaussian, with $1-2\sigma$ uncertainties shaded. Extrapolating the evidence increments in a linear/exponential manner under/over-predicts the endpoint, both of which do considerably worse than the method of extrapolating the likelihood.}
\label{fig:inc_predictions}
\end{figure}

\section{Results}\label{sec:results}
Now that we have a method for predicting the endpoint, it can be tested on a range of distributions. We begin by considering a series of toy examples to explore the capabilities and limitations of the method, before presenting results for real cosmological chains.
\subsection{Toy examples}
\subsubsection{Gaussians}
Predictions for spherical Gaussians of various dimensions are shown in \cref{fig:gauss_predictions} as a benchmark for when fitting a Gaussian distribution is exact. In all endpoint prediction plots, the shading indicates the $1-2\sigma$ uncertainties. All were run with $\nlive = 500$ except for one $\nlive=2000$ for comparison, with each Gaussian having a width of $\sigma = 0.01$. The correct endpoint is recovered to within standard error at all points except the very beginning, when the parameters have hardly been constrained. 
\par
We note that as with nested sampling in general, increasing $\nlive$ improves the resolution and reliability of the inferences, which can be seen from the middle two plots.
\begin{figure*}
\begin{center}
    \includegraphics{figures/gauss_predictions.pdf}
\end{center}
\caption{Endpoint predictions for a spherical Gaussian run with $\nlive=500$ (except for the third plot from left). The correct endpoint is obtained for all but the earliest iterations, and the uncertainty is controlled by the number of live points, which can be seen from the two $d = 16$ plots.}
\label{fig:gauss_predictions}
\end{figure*}
We also observe the effect of elongating the Gaussian, using the same example as \cref{sec:elongated}. \cref{fig:elongated_logXfs} shows a step-like trend similar to the inferred dimensionalities, reflecting the fact that the full dimensionality is undetectable at lower compression factors. The endpoint for a likelihood whose remaining three directions are completely unconstrained coincides with our predictions at early iterations, showing that the two cases are indistinguishable. 
\begin{figure*}
\begin{center}
    \includegraphics{figures/elongated_logXfs}
\end{center}
\caption{Endpoint prediction for an elongated Gaussian. At early stages, the full dimensionality is undetectable, and the endpoint is predicted to be the same as for a likelihood with three unconstrained directions. Only once the prior has been compressed enough to constrain the other three directions does the prediction converge to the true value.}
\label{fig:elongated_logXfs}
\end{figure*}

\subsubsection{Cauchy}
One case that might be expected to cause problems is the pathological Cauchy distribution, which is far from a Gaussian. \cref{fig:cauchy_predictions} shows the predictions for a likelihood of the form
\begin{equation}
	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{1 + d}{2} \log \left(1 + \frac{X^2}{\gamma^2}\right),
\end{equation}
choosing $d = 10$ and allowing $\gamma$ to vary. The correct estimate is obtained to within standard error by about halfway, but before that is quite off. The key limitation is that the estimate is wrong early on not because the compression is anisotropic, or that there is a phase transition; rather, it is a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
\par
Nonetheless, the right order of magnitude is obtained at all times, so this is not a major concern. The Cauchy is also a pathological case, and the same problem does not rear its head in more realistic cases, as we shall see next.
\begin{figure*}
\begin{center}
    \includegraphics{figures/cauchy_predictions.pdf}
\end{center}
\caption{Predictions for the Cauchy distribution for various widths $\gamma$ and $\nlive = 500$. The endpoint is underestimated for the first half of the run, but this is a limitation of the Gaussian approximation rather than a lack of information mid-run.}

@@

\section{Conclusion}
We have derived new analytic results to make an anatomy of nested sampling, understanding the progression of a run via the compression of prior volume, the increase in log-likelihood, the inferred temperature schedule, and the convergence of the sample dimensionality. From these analyses, we developed a method for predicting the endpoint of a nested sampling run, by using the inferred Bayesian model dimensionality mid-run to extrapolate the known likelihood profile. 
\par
The method in general converges to the correct endpoint by about the halfway point, and gets the correct order of magnitude throughout. Consistent predictions are obtained for both toy and cosmological examples. The accuracy is typically limited by the information available mid-run, either because of a phase transition or because the anisotropy of the nested sampling compression. Pathological distributions, such as a Cauchy, lead to less stable inferences of the dimensionality and expose the limitations of a Gaussian approximation, though the order of magnitude is still correct.
\par
Further work can be done to experiment with more flexible basis functions for regression of the likelihood profile, so that it is less dependent on the Gaussian approximation.

\section{Code availability}
A package implementing the endpoint prediction mechanism is in development for easy plug-in to existing implementations of nested sampling. The latest updates can be found at \href{https://github.com/zixiao-h/aeons}{github.com/zixiao-h/aeons}.

>>>>

\begin{abstract}
    This paper presents analytic results on the anatomy of nested sampling, from which a technique is developed to estimate the run-time of the algorithm that works for any nested sampling implementation. We test these methods on both toy models and true cosmological nested sampling runs. The method gives an order-of-magnitude prediction of the end point at all times, forecasting the true endpoint within standard error generally around the halfway point.  
\end{abstract}

@@

\subsubsection*{Comparisons}
\cref{fig:beta_comparison} shows the three temperatures as a function of compression for two cases, one containing a phase transition and one without. They are consistent in both cases when there is a single dominant phase, but differ during a phase transition. The canonical temperature is the only one that rises monotonically with compression.
\begin{figure}
\begin{center}
    \includegraphics{figures/beta_comparison.pdf}
\end{center}
\caption{Inferred temperatures using the microcanonical, canonical and Bayesian definitions. The shaded regions show the $1-2\sigma$ uncertainties. All are consistent for a single phase, but differ during a phase transition.}
\label{fig:beta_comparison}
\end{figure}
\par
One should keep in mind that despite the above theoretical reasoning, our introduction of the likelihood transformation was ultimately motivated by our wish to utilise the extra degree of freedom it provides. As we will see below, we recommend choosing the exact definition depending on what is useful for the problem at hand.

@@

\section{Endpoint prediction}\label{sec:endpoint}
As described in \citet{supernest} and further explored in the talk and upcoming paper \citet{kcl_talk, scaling_frontier}, the time complexity of nested sampling is
\begin{equation}
    T \propto \langle \tfrac{1}{\nlive} \rangle^{-1} \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
\end{equation}
The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n)$, The second term is the average time per likelihood evaluation. The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude. 
The final term is the Kullback-Liebler divergence~\cref{eq:DKL}, the compression factor required to get from the prior to the posterior. This term is generally outside of user control, in most cases \textit{a priori} unknown, and of principle interest in this section.

\subsection{The termination prior volume}
Making the above discussion more precise, we wish to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}). The difficulty is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is far from the posterior bulk. 
\par
In order to get an idea of where the true posterior bulk sits, we need to predict what the posterior looks like past the highest live point. We do this by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points. 
One would never use this predicted posterior to perform inference, since more accuracy can always be achieved by simply finishing the run. However, we will demonstrate it is sufficient for making a run-time prediction for $\log X_\mathrm{f}$. 
\par
Quantitatively, this proceeds as follows: fit a function $f(X, \phi)$  with some parameters $\phi$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
\begin{equation}
	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
\end{equation}
or equivalently
\begin{equation}\label{endpoint}
	\int_0^{X_\mathrm{f}} f(X, \phi)\ \mathrm{d}X = \epsilon \left( \int_0^{X_i} f(X, \phi)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_\mathrm{f}$ can then be identified by solving the above equation either analytically or numerically. 
\par
Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule. For example in the constant $\nlive$ case at each iteration $\log X$ decreases by $1/n$, so the total number of iterations $N_\mathrm{f}$ would be
\begin{equation}
	N_\mathrm{f} = - n\log X_\mathrm{f} .
\end{equation}

\subsection{How to extrapolate?}
A key observation is that the Bayesian model dimensionality is the equivalent dimension of the posterior if it were actually Gaussian. Fitting a Gaussian of this dimension to the likelihood profile therefore makes a reasonable approximation to the true distribution, without explicitly assuming the form of the likelihood function. The parameterisation of the Gaussian that we fit is the same as that given in \cref{sec:logL}, which we shall repeat here for clarity;
\begin{equation}\label{gaussian}
    f(X; \phi) = \logLmax - X^{2/d}/2\sigma^2
\end{equation}
The extrapolation then proceeds thus:
\begin{enumerate}[leftmargin=*]
    \item Find the current dimensionality ${d}^{(i)}_G$ of the posterior at the Bayesian temperature
    \item Take the live point profile and perform a least squares fit to \eqref{gaussian}, stipulating that $d = {d}^{(i)}_G$ to infer $\logLmax$ and $\sigma$ 
    \item Use the likelihood predicted by these parameters to solve \eqref{endpoint} for $X_\mathrm{f}$
\end{enumerate}
The advantage of fitting a Gaussian is that the procedure can be sped up analytically. Firstly, the least squares regression is trivial because analytic estimators exist; the cost function 
\begin{equation}\label{chi squared}
	C^2(\logLmax, \sigma) = \sum_i \left| \log \Like_i - f(X_i; \logLmax, \sigma) \right| ^2
\end{equation}
is minimised with respect to $(\logLmax, \sigma)$ when
\begin{equation}\label{eq:sigma}
    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i X_i^{2/d}\log \Like_i },
\end{equation}
and
\begin{equation}\label{eq:logLmax}
    \logLmax = \frac{1}{N} \sum_i \log \mathcal{L}_i + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d}.
\end{equation}
Secondly, the termination prior volume can also be obtained analytically. Rewriting \cref{endpoint} in terms of the Gaussian parameters gives
\begin{equation}
	\epsilon = \frac{\int_0^{X_\mathrm{f}} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead}}.
\end{equation}
The integrals have the analytic solution
\begin{equation}
	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X = \frac{d}{2} \cdot \left(\sqrt{2}\sigma\right)^d \cdot \gamma_k
\end{equation}
where $\gamma_k = \Gamma_{d/2}\left(X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
\begin{equation}
    \log X_\mathrm{f} = \frac{d}{2}\log 2\sigma^2	+ \log \Gamma^{-1}_{d/2} \left(\epsilon \gamma_i+ \frac{\epsilon\mathcal{Z}_\mathrm{dead}}{ \left( 2\sigma^2 \right)^{d/2}\Like_\mathrm{max}}\right),\label{eq:xf}
\end{equation}
and $N_\mathrm{f}$ is of course just $-n$ multiplied by this. Intuitively, the above procedure can be thought of as inferring the number of constrained parameters, then extrapolating them up to find the point at which they will be fully constrained. 
\par
One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown this approach to be far less stable.

\subsection{Alternative approaches}
More comprehensive Bayesian approaches, perhaps including \textit{a priori} information about the likelihood or greater flexibility in the fitting function, could likely perform better than what we have just presented. However, such methods would not befit \textit{run-time} prediction which has a much more limited computational budget, hence the more pragmatic approach we have adopted. Here, we discuss as a benchmark alternative approaches to endpoint estimation that have a comparable computational complexity. 
\subsubsection{Integral progress}
An alternative approach used in \textsc{Ultranest} \citep{ultranest} derives a progress bar based on the fraction of the accumulated integral compared to the remaining integral, approximated as 
\begin{equation}\label{eq:integral_progress}
    \mathcal{Z}_\mathrm{rem} \approx \Like_\mathrm{max}^{\mathrm{live}} X_i.
\end{equation}
This has a several shortcomings. First, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width. Second, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is almost zero for most of the run. Finally, approximation \eqref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum. This approach can however be useful in the low dimensions appropriate for \textsc{Ultranest} when the live points are always near the maximum, but in general is less reliable.

\subsubsection{Extrapolating evidence increments}
Seasoned watchers of nested sampling runs might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges. We do this for a spherical Gaussian and compare it to our method. At an intermediate stage of the run, the most recent outputs might look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate}. Extrapolating those data to a linear and exponential profile yields endpoint estimates plotted in the graph to the right. 
\par
The linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile. The increments are also not exactly exponential, since the exponential fit leads to a large over-prediction. The predicted endpoint over the course of a run for $d = 16$, $\sigma = 0.01$, as shown in \cref{fig:inc_predictions}, shows the same result. One might expect an average to be more accurate, but this tends to be biased towards the exponential prediction, and there is no obvious choice of weighting that would fix this.
\par
More importantly, we find that for real likelihoods which have an element of noise the extrapolation often diverges, for instance when the increments do not monotonically decrease. Directly extrapolating the evidence increments is therefore far less stable than the previous method, and generally not a reliable method for prediction.
\begin{figure}
\begin{adjustbox}{valign=t, scale=0.72}
\subfloat{\begin{tabular}{|c|c|c|}
\hline
iteration & log Z  & $\Delta\log Z$   \\
\hline
5000 & -1435.8 & 190.8 \\
5500 & -1264.6 & 171.2 \\
6000 & -1123.7 & 140.9 \\
6500 & -991.5 & 132.2 \\
7000 & -885.0 & 106.6 \\
7500 & -790.3 & 94.7 \\
8000 & -702.6 & 87.7 \\
8500 & -619.7 & 82.9 \\
9000 & -551.8 & 67.9 \\
9500 & -492.7 & 59.1 \\
\hline
\end{tabular}}
\end{adjustbox}
\quad
\begin{adjustbox}{valign=t}
\subfloat{\includegraphics{figures/inc_extrapolate.pdf}}
\end{adjustbox}
\caption{Extrapolating the increments of evidence. The left column shows the output of a nested sampling run, and the right column shows the extrapolation.}
\label{fig:inc_extrapolate}
\end{figure}
\begin{figure}
\includegraphics{figures/inc_predictions.pdf}
\caption{Endpoint predictions for a spherical Gaussian, with $1-2\sigma$ uncertainties shaded. Extrapolating the evidence increments in a linear/exponential manner under/over-predicts the endpoint, both of which perform considerably worse than the method of extrapolating the likelihood.}
\label{fig:inc_predictions}
\end{figure}

\section{Results}\label{sec:results}
We now test the approach established in the previous section on a range of distributions. We begin by considering a series of toy examples to explore the capabilities and limitations of the method, before presenting results for real cosmological chains.
\subsection{Toy examples}
Throughout the toy examples we make use the perfect nested sampling~\citep{Keeton_2011,2018BayAn..13..873H} framework implemented in \textsc{anesthetic}~\cite{anesthetic}.
\subsubsection{Gaussians}
Predictions for spherical Gaussians of various dimensions are shown in \cref{fig:gauss_predictions} as a benchmark for when fitting a Gaussian distribution is exact. In all endpoint prediction plots, the shading indicates the $1-2\sigma$ uncertainties. All were run with $\nlive = 500$ except for one $\nlive=2000$ for comparison, with each Gaussian having a width of $\sigma = 0.01$. The correct endpoint is recovered to within standard error at all points except the very beginning, when the parameters have hardly been constrained. 
\par
We note that as with nested sampling in general, increasing $\nlive$ improves the resolution and reliability of the inferences, which can be seen from the middle two plots.
\begin{figure*}
\begin{center}
    \includegraphics{figures/gauss_predictions.pdf}
\end{center}
\caption{Endpoint predictions for a spherical Gaussian run with $\nlive=500$ (except for the third plot from left). The correct endpoint is obtained for all but the earliest iterations, and the uncertainty is controlled by the number of live points, which can be seen from the two $d = 16$ plots.}
\label{fig:gauss_predictions}
\end{figure*}
We also observe the effect of elongating the Gaussian, using the same example as \cref{sec:elongated}. \cref{fig:elongated_logXfs} shows a step-like trend similar to the inferred dimensionalities, reflecting the fact that the full dimensionality is undetectable at lower compression factors. The endpoint for a likelihood whose remaining three directions are completely unconstrained coincides with our predictions at early iterations, showing that the two cases are indistinguishable. 
\begin{figure}
\begin{center}
    \includegraphics{figures/elongated_logXfs}
\end{center}
\caption{Endpoint prediction for an elongated Gaussian. At early stages, the full dimensionality is undetectable, and the endpoint is predicted to be the same as for a likelihood with three unconstrained directions. Only once the prior has been compressed enough to constrain the other three directions does the prediction converge to the true value.}
\label{fig:elongated_logXfs}
\end{figure}

\subsubsection{Cauchy}
One case that might be expected to cause problems is the pathological Cauchy distribution, which is far from a Gaussian. \cref{fig:cauchy_predictions} shows the predictions for a likelihood of the form
\begin{equation}
	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{1 + d}{2} \log \left(1 + \frac{X^2}{\gamma^2}\right),
\end{equation}
choosing $d = 10$ and allowing $\gamma$ to vary. The correct estimate is obtained to within standard error by about halfway, but before that is inaccurate. The key limitation is that the estimate is wrong early on, not because the compression is anisotropic, or that there is a phase transition; but rather as a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
\par
Nevertheless, the right order of magnitude is obtained at all times, so this remains sufficient for most use-cases. The Cauchy is also a pathological case, and the same problem does not in practice appear for more realistic cases, as we shall see next.
\begin{figure*}
\begin{center}
    \includegraphics{figures/cauchy_predictions.pdf}
\end{center}
\caption{Predictions for the Cauchy distribution for various widths $\gamma$ and $\nlive = 500$. The endpoint is underestimated for the first half of the run, but this is a limitation of the Gaussian approximation rather than a lack of information mid-run.}

@@

\section{Conclusion}
We have derived new analytic results to make an anatomy of nested sampling, understanding the progression of a run via the compression of prior volume, the increase in log-likelihood, the inferred temperature schedule, and the convergence of the sample dimensionality. From these analyses, we developed a method for predicting the endpoint of a nested sampling run, by using the inferred Bayesian model dimensionality mid-run to extrapolate the known likelihood profile. 
\par
The method in general converges on a correct prediction of endpoint by about halfway, and gets the correct order of magnitude throughout. Consistent predictions are obtained for both toy and cosmological examples. The accuracy is typically limited by the information available mid-run, either because of a phase transition or because the anisotropy of the nested sampling compression. Pathological distributions, such as a Cauchy, lead to less stable inferences of the dimensionality and expose the limitations of a Gaussian approximation, though the order of magnitude is still correct.
\par
Further work can be done to experiment with more flexible basis functions for regression of the likelihood profile, so that it is less dependent on the Gaussian approximation.

\section*{Code availability}
A package implementing the endpoint prediction mechanism is in development for easy plug-in to existing implementations of nested sampling. The latest updates can be found at \href{https://github.com/zixiao-h/aeons}{github.com/zixiao-h/aeons}.

\section*{Acknowledgements}
WJH is supported by the Royal Society as a Royal Society University Research Fellow at the University of Cambridge. AB was supported by a Cambridge Mathematical Placement and the Royal Society summer studentship. ZH was supported by a Royal Society summer studentship. We thank Mike Hobson for helpful discussion during ZH Part III viva.
